---
title: "Getting the Most Out of Functions"
author: "Steve Pederson"
date: "4 December 2015"
output: ioslides_presentation
---

## Functions - An Introduction

- Functions are the "engine"" of R

- We use them to "do things", then we obtain some output

- Some well known ones are `max()`, `min()`, `mean()`, `sd()`

```{r, eval=FALSE}
y <- c(1, 16, 2, 9)
mean(y)
min(y)
sd(y)
sqrt(y)
```

## Functions - An Introduction

- Usually take some input, but some special functions don't

```{r, eval=FALSE}
browseVignettes()
help.start()
Sys.Date()
```

- To execute a function, we need to place round brackets `()` after the name  
(Otherwise we see the code)

**An R Package is essentially just a collection of functions!**


## Why would we write functions?

- Repetitive processes

- Nothing may exist for our purpose

- We can customise existing function
    + We may need to tweak the output of a function
    + We may only need part of a larger function
    + We can customise for increased speed

## Our first function

```{r}
myMean <- function(x){
  n <- length(x)
  total <- sum(x)
  return(total / n)
}
```

The functions needs:

1. A `name` so we can 'call' it
2. An `argument` (or input) to 'do something' to.  
   Known in *R* world as the `formals`
3. A `body` which is the code that runs when we call the function

## Our first function

```{r}
myMean <- function(x){
  n <- length(x)
  total <- sum(x)
  return(total / n)
}
```

**NB**

- `x` is just a place holder for "some R object"

- This object then becomes called `x` as far as the function is concerned

## Our first function

Let's run it:

```{r}
y <- c(1, 16, 2, 9)
myMean(y)
```

Check it against the real function

```{r}
mean(y)
```

**NB:** We didn't need to write `myMean(x = y)`

## Our first function

To learn about a function, we can see what arguments are required
```{r}
formals(myMean)
```

We can inspect the code
```{r}
body(myMean)
```

## Help pages

We can read the help page (if one exists)
```{r, eval=FALSE}
?median
?sd
?mean
```

A well written help page is amazing

(These can be rare for beginners in R)

## Writing functions for NOT you

(I mean you in 6 months, when you have no idea what you did)

We can place comments in the code

```{r}
myMean <- function(x){
  
  # First, count how many values are in x
  n <- length(x)
  
  # Now add them
  total <- sum(x)
  
  # And produce some nice output
  total / n
  
}
```

## Writing functions for NOT you

Notice that these comments are not shown when we call `body()`

```{r}
body(myMean)
```

Comments are a great idea for new to intermediate *R* users!

## What could possibly go wrong?

Now we know how to write function that works:  
**How could this ever go wrong?**

## What could possibly go wrong?

Some ways to break `myMean` are:

1. Give it `NA` values
2. Give it something which is not a `numeric` vector

```{r, echo=FALSE, include=FALSE}
ohNo <- c(NA, 1, 3, 5)
```


```{r,eval = FALSE}
ohNo <- c(NA, 1, 3, 5)
myMean(ohNo)
```

## Debugging!

- We might need to debug the function.

- R has an inbuilt function called `browser()`

- If we (temporarily) place this command inside the function, we move
    + **from the Global R Environment**  
    + **to an R Environment INSIDE the function**

## Debugging!

```{r}
myMean <- function(x){
  
  browser()
  
  # First, count how many values are in x
  n <- length(x)
  
  # Now add them
  total <- sum(x)
  
  # And produce some nice output
  total / n
  
}
```

## Debugging!

Now let's call the function, and a new window will open

You'll also see some new buttons at the top of the R Console

```{r, eval=FALSE}
myMean(ohNo)
```

Turn to the `Environment` Tab at the top right of R Studio

- This is displaying the internals of the function

## Debugging!

Send the first line in the `Source Viewer` Window to the Console

- `n` will appear in the `Environment` Tab (and so will `x`)

- The value looks right, so this is OK

Now, send the second line to the Console

- The value for `total` is not what we expected, so we need to fix this line!

**Suggestions?**

```{r, eval=FALSE}
?sum
```

## Debugging!

```{r}
myMean <- function(x){
  
  browser()
  
  # Find the values which are not NA
  notNA <- which(!is.na(x))
  
  # Find how many valid values there are
  n <- length(x[notNA])
  
  # Now add the valid numbers
  total <- sum(x[notNA])
  
  # And produce some nice output
  total / n
  
}
```

## Debugging!

My final function would look something like this:
```{r}
myMean <- function(x){
  
  # Check the vector is numeric
  stopifnot(is.numeric(x))
  
  # Count the valid values
  n <- sum(!is.na(x))
  
  # Add the valid values
  total <- sum(x, na.rm = TRUE)
  
  # And now provide some output
  total / n
  
}
```

