---
title: "Session 2 - Customising Functions"
author: "Steve Pederson"
date: "4 December 2015"
output: ioslides_presentation
---

## Digging Deeper

R comes with numerous in-built datasets

```{r}
?ToothGrowth
```

Here, the length of growing teeth has been measured using:

- 2 Vitamin C Supplements (Orange Juice or Ascorbic Acid)

- 3 different dose levels (0.5, 1.0 & 2.0mg)

## Tooth Growth

First let's 

1. make a shortcut to this dataset called `tg`

2. change the dose levels to a categorical variable (i.e a `factor`)

```{r}
tg <- ToothGrowth
tg$dose <- factor(tg$dose)
levels(tg$dose) <- c("Low", "Med", "High")
```

## Data Inspection

Firstly we should inspect and plot our data

```{r}
head(tg)
```

## Data Inspection
```{r, fig.height=4}
library(ggplot2)
ggplot(tg, aes(x = dose, y = len, fill = supp)) +
  geom_boxplot()
```


## Tooth Growth

Our task is to conduct a `t.test()` for each dose level to see if there is any effect on tooth length.

Let's try it for the low dosage level

```{r}
t.test(len ~ supp, data = tg, subset = dose == "Low")
```


## Tooth Growth

- Doing this for all 3 levels will look messy! 

- Let's change the structure of the output to make it look nicer

- First, how does R really save the results from the `t.test()` function?

```{r, eval=FALSE}
results <- t.test(len ~ supp, data = tg, subset = dose == "Low")
str(results)
```

## Tooth Growth

This is an *R* object type called a `list`

*What are we really interested in?*

## Tooth Growth

This is an *R* object type called a `list`

*What are we really interested in?*

1. The difference in means?

2. The *T*-statistic

3. The degrees of freedom

4. The *p*-value

**Where are these values in the function output?**

## Tooth Growth

We could write a function to output these as a `data.frame`

```{r}
myToothTest <- function(data){
  
  # Perform the t.test
  res <- t.test(len~supp, data)
  
  # Output the data.frame
  data.frame(Difference  = -diff(res$estimate),
             T = res$statistic,
             DF = res$parameter,
             P = res$p.value, 
             row.names= NULL)
  
}
```

## Tooth Growth

- Now we can run the function

- Note that we're going to do the sub-setting first

```{r}
subTg <- subset(tg, dose == "Low")
myToothTest(subTg)
```

- This looks much nicer!

## Tooth Growth

- Let's incorporate the sub-setting into the function

- AND add the dose level to the output

## Tooth Growth

```{r}
myToothTest <- function(doseLevel, data){
  
  stopifnot(doseLevel %in% levels(data$dose))
  data <- subset(data, dose == doseLevel)
  
  # Perform the t.test
  res <- t.test(len~supp, data)
  
  # Output the data.frame
  data.frame(Dose = doseLevel,
             Difference  = -diff(res$estimate),
             T = res$statistic,
             DF = res$parameter,
             P = res$p.value, 
             row.names= NULL)
  
}
```

## Tooth Growth

Now the function is even easier to run

```{r}
myToothTest("Low", tg)
```

## Now the FUN starts

*R* has a function called `lapply()`

This will apply a function to each element of a vector

```{r, eval=FALSE}
?lapply
```

```{r, echo=FALSE}
y <- c(1,16, 2, 9)
```

Just to see this in action:
```{r, eval=FALSE}
y <- c(1, 16, 2, 9)
lapply(y, FUN = sqrt)
```

**This places each element in the FIRST argument position of the function**

## Now the FUN starts

There is a alternate function `sapply()` that simplifies the output where possible

```{r}
sapply(y, FUN = sqrt)
```

It's very useful, but slightly unpredictable...

## Now the FUN starts

*Can we use this idea for our different dose levels?*

## Now the FUN starts

*Can we use this idea for our different dose levels?*

YES!!!

This is why I placed `doseLevels` as the first argument of our function

```{r, echo=TRUE, results='hide'}
doseLevels <- levels(tg$dose)
lapply(doseLevels, myToothTest, data = tg)
```

Note that we specified `data = tg` after the function call

## Getting amazing looking output

The *R* package `dplyr` has a function `bind_rows()` for joining identical `data.frame` objects.

First install and/or load the package
```{r, eval=FALSE}
install.packages(dplyr)
library(dplyr)
```

## Getting amazing looking output

```{r, echo=FALSE, message=FALSE}
library(dplyr)
```

```{r, warning=FALSE}
doseResults <- lapply(doseLevels, myToothTest, data = tg)
doseResults <- bind_rows(doseResults)
doseResults
```

## Getting amazing looking output

And for those using R Markdown/knitr

```{r, warning=FALSE}
library(knitr)
kable(doseResults)
```

## In the real world

This technique can be applied to numerous datasets

- You might want to run the same linear regression model on data from different climates/regions

- You might have thousands of SNPs to run a `fisher.test()` on
